Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-09 dev/paper/paper.tex
bleh

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on algorithmic recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
which enables us to study macrodynamics of algorithmic recourse through
simulations. To this end we propose a number of novel evaluation metrics
that can be used to qunatify and benchmark the macrodynmics introduced
by different counterfactual generators. Thirdly, we use this framework
to provide a first in-depth analysis of endogenous recourse dynamics
induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. Finally, we
discuss what drives endogenous dynamics and strategies to mitigate them.

@@

@@

Despite the large and growing number of approaches to counterfactual
search there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework that can be used to apply and benchmark different methodology;
finally,
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
provides an extensible and language-agnostic implementation in Julia.
For the experiments conducted in this work we relied on the latter.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic
Recourse (AR) has largely been limited to the static setting: given some
classifier we are interested in finding close, actionable, realistic,
sparse, diverse and ideally causally founded counterfactuals. The
ability of CE to handle dynamics like data and model drift remains a
largely unexplored research challenge at this point. Only one recent
work considers the implications of exogenous domain and model shifts.
This project instead focuses on endogenous dynamics, that is shifts that
occur when AR is actually implemented by a proportion of individuals.
Early findings suggest that the involved shifts may be large with
important implications on the validity of AR and the overall
characteristics of the sample population.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

To the best of our knowledge this is the first work investigating
endogenous macrodynamics in AR. Our contributions to the state of
knowledge are as follows: firstly, we posit a compelling argument that
calls for a novel perspective on algorithmic recourse extending our
focus from single individuals to groups. Secondly, we introduce an
experimental framework extending previous work by
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}},
which enables us to study macrodynamics of algorithmic recourse through
simulations that can be fully parallelized. Thirdly, we use this
framework to provide a first in-depth analysis of endogenous recourse
dynamics induced by various popular counterfactual generators including
\protect\hyperlink{ref-wachter2017counterfactual}{{[}5{]}},
\protect\hyperlink{ref-schut2021generating}{{[}7{]}},
\protect\hyperlink{ref-joshi2019towards}{{[}8{]}},
\protect\hyperlink{ref-mothilal2020explaining}{{[}9{]}} and
\protect\hyperlink{ref-antoran2020getting}{{[}10{]}}. To this end we
propose a number of novel evaluation metrics that can be used to
quantify and benchmark the macrodynamics introduced by the different
generators. Finally, we also discuss what drives endogenous dynamics and
propose strategies to mitigate them.

@@

As mentioned earlier, research has so far also generally focused on
generating counterfactuals for single individuals or instances. We have
been able to identify only one existing work that investigates black-box
model behavior towards a group of individuals
\protect\hyperlink{ref-carrizosa2021generating}{\textbf{carrizosa2021generating?}}.
The authors propose an optimization framework that generates collective
counterfactuals. We provide a motivation for doing so from the
perspective of endogenous macrodynamics of algorithmic recourse.

@@

Despite the large and growing number of approaches to counterfactual
search there have been surprisingly few benchmark studies that compare
different methodologies. This may be partially due to limited software
availability in this space. Recent work has started to address this gap:
firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large
benchmarking study using different algorithmic aproaches and numerous
tabular datasets; secondly,
\protect\hyperlink{ref-pawelczyk2021carla}{{[}4{]}} introduce a Python
framework that can be used to apply and benchmark different
methodologies; finally,
\protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}6{]}}
provides an extensible, fast and language-agnostic implementation in
Julia. Since the experiments presented here involve extensive
simulations, we have relied on and extended the Julia implementation due
to the associated performance benefits.
